Order reduction and MILP models for real-time DNN robustness certification.

### 2025-08-02 AAAI2026/Lipitschz.tex
final?

In this paper, we introduced a novel "2v" MILP model for global verification questions, which is 4 to 8 times more accurate than the classical MILP model,
and more accurate than the ITNE model of \cite{lipshitz}, see 
Table \ref{table.classical}. Using our "3v" or "1v" models abstracting away some features of the "2v" model,
 even better bounds are reached (Tables 3, 4, 5). We also explained how to support $L_1$-perturbations efficiently on top of classical $L_\infty$-perturbations (Prop \ref{prop.l1}). To further lower the bounds by $> 10$ times, we resort to model order reduction, reducing the space of search closer to the space of the training dataset, without sacrificing DNN accuracy ($97\%$). This results into, to the best of our knowledge, the first {\em real-time} robustness certification reported for a DNN, with $86\%$ of robustness for an $L_1$-perturbation of $.5$ and $53\%$ of robustness for an $L_1$-perturbation twice as large of $1$. Still, global verification is only at its beginning, and combining different methods is necessary to scale to more complex DNNs.

>>>>

In this paper, we introduced a novel "2v" MILP model for global verification questions, which is 4 to 8 times more accurate than the classical MILP model,
and more accurate than the ITNE model of \cite{lipshitz,ITNE}, see 
Table \ref{table.classical}. Using our "3v" or "1v" models abstracting away some features of the "2v" model,
 even better bounds are reached (Tables 3, 4, 5). We also explained how to support $L_1$-perturbations efficiently on top of classical $L_\infty$-perturbations (Prop \ref{prop.l1}). To further lower the bounds by $> 10$ times, we resort to model order reduction, reducing the space of search closer to the space of the training dataset, without sacrificing DNN accuracy ($97\%$). This results into, to the best of our knowledge, the first {\em real-time} robustness certification reported for a DNN, with $86\%$ of robustness for an $L_1$-perturbation of $.5$ and $53\%$ of robustness for an $L_1$-perturbation twice as large of $1$. Still, global verification is only at its beginning, and combining different methods is necessary to scale to more complex DNNs.

### 2025-08-02 AAAI2026/experiments.tex
final?

	We consider two DNNs: one for MNIST, namely a pre-trained $5\times 100$ (available at \url{https://github.com/eth-sri/eran}), a 5 fully connected layers DNN, with 100 neurons each,	trained with diff AI \cite{deeppoly}. 
	Although it is relatively small, it is also of good quality with $97\%$ accuracy on classification, but it is not easy to certify for global robustness (the size of a network can be deceptive to know whether they are easy or hard to verify \cite{ATVA25}).  
	The second one is a 2 fully connected layers, 50 neurons each, learnt as a surrogate model, with 10 reduced input and 26 reduced output, that predictS the plastic strain in 3000 points of a mesh over a pipe given the deformation of each of this 3000 points (through a PCA order reduction \cite{Paco} encoding all 3000 points with 10 dimensions in input and 26 dimensions in output), from \cite{aiware}.

@@

We start by evaluating how accurate the 2v model is vs the classical encoding vs the interleaving twin-network encoding (ITNE) from \cite{lipshitz}, that we reimplement in Gurobi by considering the classical encoding plus the explicit constraints $\frac{x_i-x_i' - \gamma}{2} \leq \hat{x}_i-\hat{x}_i' \leq \frac{x_i-x_i' + \gamma}{2}$, corresponding to Eq. (3) of \cite{ITNE} (and also to our (\ref{eq.lpr})). We compare when half the ReLUs are approximated with the LP relaxation, and when all the ReLUs are encoded exactly. We use the pipe models which is easier to verify, and set the time-out at 1000s for all models.

@@

On the full space, the best bounds $(\beta^{.5}_{i,j})_{i < j \leq 10}$ reached are too pessimistic: no image can be certified robust in real-time. One of the reason, besides that the verification question is very complicated with 1000 binary variables for the accurate model, is that working in the full $784$ dimensional space allows for very unlikely inputs, far away from what the DNN has learnt from, making the search very long and the bound very pessimistic. As an illustration of that, we display on Fig.~\ref{fig3} the worst-case as found by our models, which is very improbable to be ever encountered in the MNIST benchmark.

>>>>

	We consider two DNNs: one for MNIST, namely a pre-trained $5\times 100$ (available at \url{https://github.com/eth-sri/eran}), a 5 fully connected layers DNN, with 100 neurons each,	trained with diffAI \cite{deeppoly}. 
	Although it is relatively small, it is also of good quality with $97\%$ accuracy on classification, but it is not easy to certify for global robustness (the size of a network can be deceptive to know whether they are easy or hard to verify \cite{ATVA25}).  
	The second one is a 2 fully connected layers, 50 neurons each, learnt as a surrogate model, with 10 reduced input and 26 reduced output, that predicts the plastic strain in 3000 points of a mesh over a pipe given the deformation of each of this 3000 points (through a PCA order reduction \cite{Paco} encoding all 3000 points with 10 dimensions in input and 26 dimensions in output), from \cite{aiware}.

@@

We start by evaluating how accurate the 2v model is vs the classical encoding vs the interleaving twin-network encoding (ITNE) from \cite{lipshitz}, that we reimplement by considering the classical encoding plus the explicit constraints $\frac{x_i-x_i' - \gamma}{2} \leq \hat{x}_i-\hat{x}_i' \leq \frac{x_i-x_i' + \gamma}{2}$, corresponding to Eq. (3) of \cite{ITNE} (and also to our (\ref{eq.lpr})). We compare when half the ReLUs are approximated with the LP relaxation, and when all the ReLUs are encoded exactly. We consider the pipe surrogate as it is easier to verify, and set the time-out at 1000s for all models.

@@

On the full space, the best bounds $(\beta^{.5}_{i,j})_{i < j \leq 10}$ reached are too pessimistic: no image can be certified robust in real-time. One of the reason, besides that the verification question is very complicated with 1000 binary variables for the accurate model, is that working in the full $784$ dimensional space allows for very unlikely inputs, far away from the training dataset, making the search space overly large. As an illustration of that, we display on Fig.~\ref{fig3} the worst-case as found by our models, which is very improbable to be ever encountered in the MNIST benchmark.

### 2025-08-02 AAAI2026/global.tex
final?

	We call this encoding $\mathcal{M}^{classical}$.
	It contains twice as many (binary) variables as for local robustness. 
	%As the worst-case complexity of solving an MILP model is exponential in the number of binary variables, it will be computationally intractable to compute $\beta^\varepsilon_{i,j}$ in this way, but for very small DNNs. 
	As explained in the previous section, the standard way to make the model more efficient is to rely on Linearly relaxing some of the binary variables into linear variables.

@@

	We now present our novel faithful "2v" model.
    First, we consider variables $x'_j,\hat{x}'_j, y_j, \hat{y}_j$, where $y_j, \hat{y}_j$ are the {\em diff variables}, with     $y_j = x_j - x'_j$ and $\hat{y}_j= \hat{x}_j - \hat{x}'_j$, as already proposed in \cite{ITNE}. 	Intuitively, the diff variables will be small as they will be carrying the difference between both the image $I$ and its perturbation $I'$.
    We will use the classical MILP encoding for $\hat{x}'_j$.

    However, unlike the Interleaving Twin-Network Encoding (ITNE) \cite{ITNE} which relies on the classical MILP encoding (\ref{eq11}) \cite{MILP} to compute $\hat{y}_j$ from the difference between 
	$\ReLU(x'_j + y_j) = \ReLU(x_j) = \hat{x_j}$
	and $\hat{x}'_j$, we compute $\hat{y}_j$ directly, without resorting to the classical encoding. To do so, we study the function of 
    $\hat{y}_j$ given $y_j$ and $x'_j$, and encode it into a novel MILP model.
    A first important observation is that $I,I'$ being perfectly symmetrical,
    assuming that $\gamma_i$ is an upper bound of $y_i$, then $-\gamma_i$ is a lower bound for $y_i$.

@@

	Besides doing LP relaxation on some of the binary variables, we propose a 
    variant of the "2v" model: the binary variables $a'$ appearing in 
    Prop. \ref{Prop2} can be decoupled from the binary variable appearing in the classical encoding of $x'_i = \ReLU(x'_i)$. While this means having 3 binary variables per unstable ReLU (which warrants the name "3v" model), the fact that these variables are decouples remove a lot of constraints in the MILP encoding, and actually helps the solver. The decoupling also means that the "3v" model is not as accurate as the "2v" model.

>>>>

	We call this encoding $\mathcal{M}^{classical}$.
	It contains twice as many (binary) variables as for local robustness. 
	%As the worst-case complexity of solving an MILP model is exponential in the number of binary variables, it will be computationally intractable to compute $\beta^\varepsilon_{i,j}$ in this way, but for very small DNNs. 
	As explained in the previous section, the standard way to make the model more efficient is to rely on 
	linearly relaxing some of the binary variables into linear variables.

@@

	We now present our novel faithful "2v" model.
    First, we consider variables $x'_j,\hat{x}'_j, y_j, \hat{y}_j$, where $y_j, \hat{y}_j$ are the {\em diff variables}, with     $y_j = x_j - x'_j$ and $\hat{y}_j= \hat{x}_j - \hat{x}'_j$, as already proposed in \cite{lipshitz,ITNE}. 	Intuitively, the diff variables will be small as they will be carrying the difference between both the image $I$ and its perturbation $I'$.
    We will use the classical MILP encoding for $\hat{x}'_j$.

    However, unlike the Interleaving Twin-Network Encoding (ITNE) of \cite{lipshitz,ITNE} which relies on the classical MILP encoding (\ref{eq11}) \cite{MILP} to compute $\hat{y}_j$ from the difference between 
	$\ReLU(x'_j + y_j) = \ReLU(x_j) = \hat{x_j}$
	and $\hat{x}'_j$, we compute $\hat{y}_j$ directly, without resorting to the classical encoding. To do so, we study the function of 
    $\hat{y}_j$ given $y_j$ and $x'_j$, and encode it into a novel MILP model.
    A first important observation is that $I,I'$ being perfectly symmetrical,
    assuming that $\gamma_i$ is an upper bound of $y_i$, then $-\gamma_i$ is a lower bound for $y_i$.

@@

	Besides doing LP relaxation on some of the binary variables, we propose a 
    variant of the "2v" model: the binary variables $a'$ appearing in 
    Prop. \ref{Prop2} can be decoupled from the binary variable appearing in the classical encoding of $x'_i = \ReLU(x'_i)$. While this means having 3 binary variables per unstable ReLU (which warrants the name "3v" model), the fact that these variables are decoupled removes a lot of constraints in the MILP encoding, and actually helps the solver. The decoupling also means that the "3v" model is not as accurate as the "2v" model.

### 2025-08-02 AAAI2026/related.tex
final?

Closer to our paper, several works  \cite{Marabou,lipshitz,GROCET} consider a global certified restricted Lipschitz bound, to understand how the output change with input perturbations. However, they do not consider whether the classification changes.

@@

Technically, none of the MILP base work \cite{vhagar,ITNE} 
consider $L_1$-perturbations. Further, they use the classical MILP encoding from \cite{MILP}, and not our novel "2v" model, nor even our "1v" model. However, \cite{ITNE} 
defines the {\em diff variables} $y,\hat{y}$, adding explicitly the linear relaxation $\frac{y_i-\gamma_i}{2} \leq \hat{y}_i \leq \frac{y_i+\gamma_i}{2}$ implied by our models, Eq. (3) in \cite{ITNE} corresponding essentially to it. \cite{vhagar} does things differently, adding per neuron constraints (depending on the perturbation) between the (binary) variables to simplify the MILP problem, which is well-adapted to occlusion and patches but less to $L_1$-perturbations.

>>>>

Closer to our paper, several works  \cite{Marabou,lipshitz,ITNE,GROCET} consider a global certified restricted Lipschitz bound, to understand how the output change with input perturbations. However, they do not consider whether the classification changes.

@@

Technically, none of the MILP base work \cite{vhagar,lipshitz,ITNE} 
consider $L_1$-perturbations. Further, they use the classical MILP encoding from \cite{MILP}, and not our novel "2v" model, nor even our "1v" model. However, \cite{ITNE} 
defines the {\em diff variables} $y,\hat{y}$, adding explicitly the linear relaxation $\frac{y_i-\gamma_i}{2} \leq \hat{y}_i \leq \frac{y_i+\gamma_i}{2}$ implied by our models, Eq. (3) in \cite{ITNE} corresponding essentially to it. \cite{vhagar} does things differently, adding per neuron constraints (depending on the perturbation) between the (binary) variables to simplify the MILP problem, which is well-adapted to occlusion and patches but less to $L_1$-perturbations.
